Investigating the Impact of Macroeconomic Indicators on the FTSE 250 Index Using Statistical and Machine Learning Methods

### 2024-08-25 tex/paper.tex
final adjustments

Macroeconomic indicators can be effectively used in machine learning 
methods to forecast stock prices by serving as input features that capture 
the economic environment’s influence on market behaviour. By carefully 
selecting the most relevant indicators and preprocessing data, 
as well as choosing appropriate models, machine learning can enhance the accuracy of stock price forecasts (\citeboth{Prasad2023}).

@@

The remainder of this paper will be organised as follows. 
Section \ref{sec:lit} provides a brief review of the literature in this problem space, and explains the rationale behind the selection of macroeconomic variables.
Section \ref{sec:meth} presents the methodology of the paper. Section 4 discusses the results and Section 5 concludes
the study.

@@

\subsection{Data}
Monthly FTSE 250 Index price data between Oct 1992 to March 2024 has been utilised for this study giving us a dataset with 379 observations.

\begin{table}[h!]
    \centering
    \caption{Data sources}
    \begin{tabular}{lll}
        \toprule
        \textbf{Variable} & \textbf{Notation} & \textbf{Source} \\
        \midrule
        FTSE 250 Index & FTSE$_{250}$ & Yahoo Finance \\
        Consumer Price Index & CPI & Office for National Statistics (ONS) \\
        USD/GBP Exchange Rate (£) & EXCHG & Federal Reserve of St Louis \\
        Interest Rate ($\%$) & INT & Bank of England \\
        Money Supply (M3) (millions £) & M3 & Bank of England \\
        \bottomrule
    \end{tabular}
\end{table}

For the interest rate data, it was necessary to address an issue raised by the fact that the Bank of England’s Monetary Policy Committee (MPC) 
does not convene on a monthly basis, and are not obligated to 
announce a rate on a regular basis. To address the resulting irregularities the data was adjusted using \ref{alg:interest_rate_adjustment}.

\begin{algorithm}[H]
    \caption{Calculate monthly interest rate}
    \label{alg:interest_rate_adjustment}
    \KwData{Interest rate data adjustments by the Bank of England’s MPC}
    \KwResult{Monthly interest rate data}

    \For{each month $m$}{
        \If{$\forall d \in D_m \; \text{(interest rate unchanged)}$}{
            \textbf{Set:} $\text{Rate}_m \leftarrow \text{Rate}_{m-1}$\;
        }
        \Else{
            \For{each day $d_j$ with a rate change}{
                \textbf{Compute:} 
                $W_{int} = \sum_{i_k \in I_m} \frac{i_k  d_{j_k}}{x}$\;
                where $I_m = \{i_1, i_2, \ldots, i_n\}$ is the set of interest rates on days with changes, 
                and $d_{j_k} \leq x$ is the number of days into the month each rate $i_k$ occurs on.
            }
            \textbf{Set:} $\text{Rate}_m \leftarrow W_{int}$\;
        }
    }
\end{algorithm}

@@

Radial basis function neural networks are a type of artificial 
neural network that use radial basis functions at each node in the 
activation layer of the network. They are able to identify complex, 
non-linear relationships between time-series variables and have been used
extensively in this aspect (\ref{subsubsec: fore}). The advantages of radial basis function neural networks over other 
neural networks in time-series forecasting lies in their 
ability to operate effectively on small sample sizes.

\subsubsection{Model Architecture}
The exact interpolation problem, which forms the motivation for radial basis function neural networks, 
involves mapping a $d$-dimensional input space $\boldsymbol{\mathrm{X}}$ to
a $c$-dimensional target space $\boldsymbol{\mathrm{y}}$. The dataset consists of $N$ input
vectors $\boldsymbol{\mathrm{x_n}}$, along with corresponding target 
values $y_n$. The objective is to find a function
  $h(\boldsymbol{\mathrm{x}})$ such that:

@@

\subsubsection{Methodology}

First, the dataset was normalised using a z-score normalisation, because neural networks work best with data
that is between a specified range (\citeboth{mendelsohn1993}).

The inputs into the network are 4-dimensional input vectors, \(\boldsymbol{\mathrm{x_i}}\),

@@

The output is a one-dimensional output vector, \(\boldsymbol{\mathrm{y_i}}\), representing differenced monthly values in the FTSE250.

@@

Following this 
it was necessary to program the radial basis function neural network into Python. The Bayesian kernel regression approach outlined by 
\citeboth{bishop1995}, using the GaussianProcessRegressor package in Python was employed to model the network. For a 
detailed description of this construction please refer to the \hyperref[sec: appendix]{Appendix}.

Based on this, the train-test process is then conducted:
\begin{enumerate}
    \item First, the optimal hyperparameters in the covariance function \eqref{eq: gimq} 
    are computed using GridSearch Cross Validation; these hyperparameters are:
    \begin{enumerate}
        \item $\alpha$: a regularisation parameter representing a small quantity
        that is added to the diagonals of the covariance matrix to ensure it is 
        invertible. 
        \item $\beta$: the power in the covariance function, which controls the shape of the function.
        \item $\sigma$: a parameter which controls how much of the noise in the data the kernel accounts for. It moderates the risk of overfitting and underfitting.
        \item Length scale: a parameter that controls the influence that each kernel is to have. Analogous to 
        the width each radial basis function neuron possesses.
    \end{enumerate}
    \item Then, training inputs $\boldsymbol{\mathrm{x}}_1, \boldsymbol{\mathrm{x}}_2,\ldots \boldsymbol{\mathrm{x}}_{n}$
    and corresponding outputs \( y = \{y_1, y_2, \dots, y_n\} \) are successively passed through the 
    covariance function $(x,x')$, to obtain the matrix of the pairwise covariance of 
    all training observations, $K$.
    \item A $4$-dimensional input space $X$ takes in an individual training observation $(\boldsymbol{\mathrm{x}}_{i}, y_i)$ and 
    calculates the posterior weights distribution. 
    Each possible weight vector $w$ represents a different function, explicitly 
    based on the radial basis function covariance function, that could fit the data. In the 
    hidden layer the expectation is then taken, which
    combines all weights and finds the mean, weighted by the 
    respective probabilities. This represents the best functional estimate of the price movement. Based on this estimate 
    the predicted output generated.
    \item The testing process is conducted similarly. The 
    posterior weight distribution that contains the information of the training phase becomes the new 
    prior distribution. From this the posterior weight distributions for the test observations \eqref{eq: testw}
    are calculated, and then the predictive output is found (\citeboth{rasmussen2006}).

@@

\section{Results}

@@

\section{Conclusion}

>>>>

Macroeconomic indicators can also be effectively used in machine learning 
methods to forecast stock prices by serving as input features that capture 
the economic environment’s influence on stock market behaviour. By carefully 
selecting the most relevant indicators and preprocessing data, 
as well as choosing appropriate models, machine learning can enhance the accuracy of stock price forecasts (\citeboth{Prasad2023}).

@@

The remainder of this paper will be organised as follows. 
Section \ref{sec:lit} provides a brief review of the literature in this problem space, and explains the rationale behind the selection of macroeconomic variables.
Section \ref{sec:meth} presents the methodology of the paper. Section \ref{sec: results} discusses the results and Section \ref{sec: conc} concludes
the study.

@@

\subsection{Data Sources}
Monthly FTSE 250 Index price data, 
monthly CPI data, monthly M3 money supply data, monthly USD/GBP exchange 
rate data and interest rate data between Oct 1992 to March 2024 has been 
utilised for this study giving a dataset with 379 observations. The sources 
are presented in Table \ref{table: tabl}.

\clearpage
\begin{table}[h!]
    \centering
    \caption{Data sources}
    \label{table: tabl}
    \begin{tabular}{lll}
        \toprule
        \textbf{Variable} & \textbf{Notation} & \textbf{Source} \\
        \midrule
        FTSE 250 Index & FTSE$_{250}$ & Yahoo Finance \\
        Consumer Price Index & CPI & Office for National Statistics (ONS) \\
        USD/GBP Exchange Rate (£) & EXCHG & Federal Reserve of St Louis \\
        Interest Rate ($\%$) & INT & Bank of England \\
        Money Supply (M3) (millions £) & M3 & Bank of England \\
        \bottomrule
    \end{tabular}
\end{table}

\subsection{Data Preprocessing}
For the interest rate data, it was necessary to address an issue raised by the fact that the Bank of England’s Monetary Policy Committee (MPC) 
does not convene on a monthly basis, and are not obligated to 
announce a rate on a regular basis. To address the resulting irregularities the data was adjusted using Algorithm \ref{alg:interest_rate_adjustment}.

\begin{algorithm}[H]
    \caption{Calculate monthly interest rate}
    \label{alg:interest_rate_adjustment}
    \KwData{Interest rate data adjustments by the Bank of England’s MPC}
    \KwResult{Monthly interest rate data (INT)}

    \For{each month $m$}{
        \If{$\forall d \in D_m \; \text{(interest rate unchanged)}$}{
            \textbf{Set:} $\text{INT}_m \leftarrow \text{INT}_{m-1}$\;
        }
        \Else{
            \For{each day $d_j$ with a rate change}{
                \textbf{Compute:} 
                $W_{int} = \sum_{i_k \in I_m} \frac{i_k  d_{j_k}}{x}$\;
                where $I_m = \{i_1, i_2, \ldots, i_n\}$ is the set of interest rates on days with changes, 
                and $d_{j_k} \leq x$ is the number of days into the month each rate $i_k$ occurs on.
            }
            \textbf{Set:} $\text{INT}_m \leftarrow W_{int}$\;
        }
    }
\end{algorithm}

The result is a dataset $X$ composed of monthly values for the FTSE250
price, CPI value, USD/GBP exchange rate, interest rate, and M3 money supply.

\subsection{Descriptive Statistics}

@@

Radial basis function neural networks are a type of artificial 
neural network that use radial basis functions at each node in the 
activation layer of the network. They are able to identify complex, 
non-linear relationships between time-series variables and have been used
extensively in this aspect. The advantages of radial basis function neural networks over other 
neural networks in time-series forecasting lies in their 
ability to operate effectively on small sample sizes.

\subsubsection{Network Architecture}
The exact interpolation problem, which forms the motivation for radial basis function neural networks, 
involves mapping a $d$-dimensional input space $\boldsymbol{\mathrm{X}}$ to
a $c$-dimensional target space $\boldsymbol{\mathrm{y}}$. The dataset consists of $N$ input
vectors $\boldsymbol{\mathrm{x_n}}$, along with corresponding target 
values $y_n$. The objective is to find a function
  $h(\boldsymbol{\mathrm{x}})$ such that:

@@

\subsubsection{Network Construction}

Neural networks work best with data that is specified between a specific 
range (\citeboth{mendelsohn1993}). Also, to achieve stationarity
it was required to take the first difference all of the variables. Therefore, all of the 
variables in the dataset described in were normalised using a z-score normalisation
and differenced, as described in Algorithm \ref{alg:z_score_differencing}.

\begin{algorithm}[H]
    \caption{Z-Score Normalisation and Differencing of Variables}
    \label{alg:z_score_differencing}
    \KwData{Dataset $X$ with variables $X_1 = FTSE_{250}, X_2 = CPI, X_3 = EXCHG, X_4 = INT, X_5 = M3$}
    \KwResult{Normalised and differenced dataset $Z$}

    \For{each variable $X_i$ in $X$}{
        \textbf{Compute:} $\mu_i \leftarrow \frac{1}{m} \sum_{j=1}^{m} X_{ij}$\;
        \textbf{Compute:} $\sigma_i \leftarrow \sqrt{\frac{1}{m} \sum_{j=1}^{m} (X_{ij} - \mu_i)^2}$\;
        \For{each observation $X_{ij}$ in $X_i$}{
            \textbf{Normalise:} $Z_{ij} \leftarrow \frac{X_{ij} - \mu_i}{\sigma_i}$\;
        }
        \For{each observation $Z_{ij}$ from the second observation to the last}{
            \textbf{Difference:} $D_{ij} \leftarrow Z_{ij} - Z_{i(j-1)}$\;
        }
    }
    \textbf{Update:} $Z_i \leftarrow D_i$ \;
\end{algorithm}

The inputs into the network are 4-dimensional input vectors, \(\boldsymbol{\mathrm{x_i}}\), from the differenced, normalised dataset $Z$

@@

The output is a one-dimensional output vector, \(\boldsymbol{\mathrm{y_i}}\) from the dataset $Z$, representing differenced monthly values in the FTSE250.

@@

The radial basis function neural network was programmed into Python. The Bayesian kernel regression approach outlined by 
\citeboth{bishop1995}, using the GaussianProcessRegressor package in Python was employed to model the network. For a 
detailed description of this construction please refer to the \hyperref[sec: appendix]{Appendix}.

Based on this, the train-test process was conducted:
\begin{enumerate}
    \item The optimal hyperparameters in the covariance function \eqref{eq: gimq} 
    are computed using GridSearch Cross Validation. These hyperparameters are:
    \begin{enumerate}
        \item $\alpha$: a regularisation parameter representing a small quantity
        that is added to the diagonals of the covariance matrix to ensure it is 
        invertible. 
        \item $\beta$: the power in the covariance function, which controls the shape of the function.
        \item $\sigma$: a parameter which controls how much of the noise in the data the kernel accounts for. It moderates the risk of overfitting and underfitting.
        \item Length scale: a parameter that controls the influence that each kernel is to have. Analogous to 
        the width each radial basis function neuron possesses.
    \end{enumerate}
    \item Training inputs $\boldsymbol{\mathrm{x}}_1, \boldsymbol{\mathrm{x}}_2,\ldots \boldsymbol{\mathrm{x}}_{n}$
    and corresponding outputs \( y = \{y_1, y_2, \dots, y_n\} \) are successively passed through the 
    covariance function $(x,x')$, to obtain the matrix of the pairwise covariance of 
    all training observations, $K$.
    \item A $4$-dimensional input space $X$ takes in an individual training observation $(\boldsymbol{\mathrm{x}}_{i}, y_i)$ and 
    calculates the posterior weights distribution. 
    Each possible weight vector $w$ represents a different function, explicitly 
    based on the radial basis function covariance function, that could fit the data. In the 
    hidden layer the expectation is then taken, which
    combines all weights and finds the mean, weighted by the 
    respective probabilities. This represents the best functional estimate of the price movement. Based on this estimate 
    the predicted output generated.
    \item The testing process was conducted similarly. The 
    posterior weight distribution that contains the information of the training phase becomes the new 
    prior distribution. From this the posterior weight distributions for the test observations \eqref{eq: testw}
    were calculated, and then the predictive output was found (\citeboth{rasmussen2006}).

@@

\section{Results}
\label{sec: results}

@@

\section{Conclusion}
\label{sec: conc}
